---
title: "Final Project Presentation"
subtitle: "How lifestyle and health factors influence sleep duration and quality"
author: "John Donnellan"
format: 
  revealjs:
    theme: "serif"
  
editor: visual
---

```{r}
library(tidyverse)
library(plotly)

sleep <- read_csv("data/Sleep_health_and_lifestyle_dataset.csv")|>
  rename(
    sleep_quality = `Quality of Sleep`,
    sleep_duration = `Sleep Duration`,
    physical_activity = `Physical Activity Level`,
    stress_level = `Stress Level`,
    bmi_category = `BMI Category`,
    sleep_disorder = `Sleep Disorder`,
    heart_rate = `Heart Rate`) |>
  select(-`Person ID`, -`Blood Pressure`, -`Daily Steps`)|>
  #Used Chat GPT to learn how to use fct_relevel and str_replace:
  #prompt: How can I adjust the response "Normal Weight" to "Normal"?
  mutate(bmi_category = str_replace(bmi_category,"Normal Weight", "Normal"),
         Occupation = str_replace(Occupation, "Sales Representative", "Salesperson"),
  #Prompt: How can I reorder facets and bars in my bar graph using factor?
         bmi_category = fct_relevel(bmi_category, "Normal", "Overweight","Obese"),
         sleep_disorder = fct_relevel(sleep_disorder, "None", "Insomnia", "Sleep Apnea")
  )
```

## Data Introduction

-   Data: Sleep Health and Lifestyle Dataset via Kaggle

-   Variables:

    -   Demographic

        -   Age, Gender, Occupation

    -   Sleep Metrics

        -   Sleep Duration, Quality of Sleep, Sleep Disorder

    -   Health Indicators

        -   BMI Category, Heart Rate, Stress Level, Physical Activity Level

## Research Question

How do lifestyle and health factors such as occupation, physical activity, stress level, BMI category, and sleep disorders affect sleep duration and sleep quality?

## Sleep vs Stress by Occupation

```{r}
by_occupation <- sleep|>
  group_by(Occupation)|>
  summarise(
    n = n(),
    avg_sleep_hrs = mean(sleep_duration),
    avg_sleep_quality = mean(sleep_quality),
    avg_stress = mean(stress_level),
    avg_activity = mean(physical_activity))

sleep_vs_stress <- ggplot(data = by_occupation, 
                          aes(x = avg_sleep_hrs, 
                              y = avg_stress, 
                              label = Occupation, 
                              color = Occupation))+
  geom_point(size = 4) +
  geom_smooth(data = sleep, 
              aes(x = sleep_duration, 
                  y = stress_level, 
                  group = 1), 
              method = lm, 
              se = FALSE, 
              color = "grey90")+
  labs(
    title = "The Effects of Stress on Sleep by Occupation",
    x = "Average Sleep Duration",
    y = "Average Stress Level",
    caption = "Each point is an Occupation"
  ) +
  theme_minimal(base_size = 20)+
  theme(legend.position = "none")+
  scale_color_brewer(palette = "Paired")
  

ggplotly(sleep_vs_stress, tooltip = "label")
  
```

## Sleep Disorders

::: panel-tabset
### General

```{r}
ggplot(data = sleep, 
       aes(x = sleep_disorder, 
           fill = sleep_disorder, 
           alpha = .7, 
           color = sleep_disorder))+
  geom_histogram(stat= "count")+
  facet_grid(~bmi_category)+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(
    title = "Sleep Disorders by BMI Catagory",
    x = "Sleep Disorder",
    y = "Count")
```

### Occupational

```{r}
sleep |>
  ggplot(aes(x = sleep_disorder,
             fill = sleep_disorder,
             color = sleep_disorder)) +
  geom_bar(alpha = 0.7) +
  facet_wrap(~Occupation) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Sleep Disorders by Occupation",
    x = "Sleep Disorder",
    y = "Count"
  )
```
:::

## Duration by BMI Catagory

```{r}
sleep |>
  ggplot(aes(x = bmi_category,
             y = sleep_duration,
             fill = bmi_category)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Sleep Duration Across BMI Categories",
    x = "BMI Category",
    y = "Sleep Duration"
  )
```

## Takeaways

Key Findings

-   Higher stress levels are associated with lower sleep duration across occupations

-   Overweight individuals report more sleep disorders, especially insomnia and sleep apnea

-   Sleep duration tends to decline as BMI increases

-   Obese individuals show the greatest variation in sleep duration

## Future Research

What causes stress in certain occupations?

Why do certain occupations have increased sleep disorder rates?
