---
title: "sleep_lifestyle"
format: html
editor: 
  markdown: 
    wrap: sentence
---

**Importing/Cleaning Data:**

```{r}
library(tidyverse)
library(plotly)

sleep <- read_csv("data/Sleep_health_and_lifestyle_dataset.csv")|>
  rename(
    sleep_quality = `Quality of Sleep`,
    sleep_duration = `Sleep Duration`,
    physical_activity = `Physical Activity Level`,
    stress_level = `Stress Level`,
    bmi_category = `BMI Category`,
    sleep_disorder = `Sleep Disorder`
  ) |>
  select(-`Person ID`, -`Blood Pressure`, - `Heart Rate`, -`Daily Steps`)|>
  #used the help of chatgpt for these mutates:
  #prompt: How can I adjust the response "Normal Weight" to "Normal"
  #Prompt: How can I reorder facets and bars in my bar graph using factor?
  mutate(bmi_category = str_replace(bmi_category,
                                       "Normal Weight", "Normal"),
         bmi_category = fct_relevel(bmi_category, "Normal", "Overweight","Obese"),
         sleep_disorder = fct_relevel(sleep_disorder, "None", "Insomnia", "Sleep Apnea")
  )
```

**Introduction:**

This dataset via Kaggle explores the relationship between lifestyle behaviors and sleep health.
It combines demographic, physiological, and behavioral variables to provide a comprehensive view of factors that may influence sleep quality and duration.

**Research Question:**

How do lifestyle factors—specifically physical activity, stress, and BMI—influence sleep quality and the likelihood of sleep disorders?

**Variables Description:**

Demographic Variables

Gender: Categorical variable indicating gender.
Age: Age of the individual (numeric).

Sleep Metrics

Sleep Duration: Number of hours an individual sleeps per night.
Quality of Sleep: A score representing sleep quality (typically on a scale, e.g., 1–10).
Sleep Disorder: Categorical variable indicating presence of a sleep disorder (e.g., None, Insomnia, Sleep Apnea).

Lifestyle Variables

Physical Activity Level: Measure of daily physical activity (often minutes per day).

Stress Level: Self-reported stress level (ordinal scale).
Occupation: Job category of the individual.

Health Indicators

BMI Category: Categorical variable (e.g., Underweight, Normal, Overweight, Obese).


Graph #1
```{r}
ggplot(sleep, aes(x = sleep_duration)) +
  geom_histogram(fill = "steelblue", aes(y = after_stat(density)))+
  geom_density(color = "steelblue", fill = "skyblue", alpha = .5)+
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep Duration",
    y = "Density"
  ) +
  theme_minimal()
```

Graph #2
```{r}
by_occupation <- sleep|>
  group_by(Occupation)|>
  summarise(
    n = n(),
    avg_sleep_hrs = mean(sleep_duration),
    avg_sleep_quality = mean(sleep_quality),
    avg_stress = mean(stress_level),
    avg_activity = mean(physical_activity))|>
  filter(n > 2)

sleep_vs_stress <- ggplot(by_occupation, aes(x = avg_sleep_hrs, y = avg_stress, label = Occupation, color = Occupation))+
  geom_point(size = 4) +
  geom_smooth(aes(x = avg_sleep_hrs, y = avg_stress), method = lm)+
  labs(
    title = "The Effects of Stress on Sleep by Occupation",
    x = "Average Sleep Duration",
    y = "Average Stress Level",
    caption = "Each point is an Occupation"
  ) +
  theme_minimal()+
  theme(legend.position = "none")
  

ggplotly(sleep_vs_stress, tooltip = "label")
  
```

Graph #3:
```{r}
ggplot(sleep, aes(x = sleep_disorder, fill = sleep_disorder, alpha = .7, color = sleep_disorder))+
  geom_histogram(stat= "count")+
  facet_grid(~bmi_category)+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(
    title = "Sleep Disorders by BMI Catagory",
    x = "Sleep Disorder",
    y = "Count")
```


Shiny App
```{r}
library(shiny)
library(tidyverse)
library(glue)

#Strictly for appearance I asked chat to help make all the variable names, title, and table labels cleaner. Definitely not necessary but I wanted it to look clean.
pretty_label <- function(x) {
  str_replace_all(x, "_", " ") |> str_to_title()
}
occupation_choices <- unique(sleep$Occupation)

#Chat Gpt helped me make the variables appear cleaner in the drop down
var_choices <- c(
  "Age"               = "Age",
  "Sleep Duration"    = "sleep_duration",
  "Sleep Quality"     = "sleep_quality",
  "Physical Activity" = "physical_activity",
  "Stress Level"      = "stress_level"
)

ui <- fluidPage(
   sidebarLayout(
    sidebarPanel(
      #Chat GPT helped me learn how to create multiple selection input
      checkboxGroupInput(
        "occupation_choose",
        label = "Select Occupation(s):",
        choices = occupation_choices,
        selected = "Doctor"),
      
      selectInput("xaxis_choose",
                  label = "Choose X-Axis Variable:",
                  choices = var_choices,
                  selected = "Age"),
      
      selectInput("yaxis_choose",
                  label = "Choose Y-Axis Variable:",
                  choices = var_choices,
                  selected = "sleep_duration")),
    
    mainPanel(
      plotOutput("final_plot"),
      plotOutput("second_plot"),
      tableOutput("final_table")
      
    )
  )
)

server <- function(input, output, session) {
  
  plot_sleep <- reactive({
    sleep |> filter(Occupation %in% input$occupation_choose)
  })
  
  output$final_plot <- renderPlot({
    ggplot(plot_sleep(),
           aes(x = .data[[input$xaxis_choose]],
               y = .data[[input$yaxis_choose]],
               color = Occupation)) +
      geom_point(size = 2) +
      theme_minimal(base_size = 15) +
      labs(
        #Chat gpt helped me learn how to use pretty_label function to clean up the   titles on graph
        title = glue("Scatterplot of {pretty_label(input$yaxis_choose)} vs {pretty_label(input$xaxis_choose)}"),
        x = pretty_label(input$xaxis_choose),
        y = pretty_label(input$yaxis_choose),
        color = "Occupation")
  })
  
  output$second_plot <- renderPlot({
    ggplot(plot_sleep(),
           aes(x = sleep_disorder,
               fill = sleep_disorder)) +
      geom_histogram(stat= "count")+
      theme_minimal(base_size = 15)+
      theme(legend.position = "none")+
      labs(
        title = "Sleep Disorders by Occupation",
        x = "Sleep Disorder",
        y = "Count")
  })
  
  output$final_table <- renderTable({
    plot_sleep() |>
      group_by(Occupation) |>
      summarise(
        N = n(),
        #Chat Gpt helped me clean up the names of the table to make it look cleaner
        !!glue("Min {pretty_label(input$yaxis_choose)}")  := min(.data[[input$yaxis_choose]]),
        !!glue("Mean {pretty_label(input$yaxis_choose)}") := mean(.data[[input$yaxis_choose]]),
        !!glue("Max {pretty_label(input$yaxis_choose)}")  := max(.data[[input$yaxis_choose]]),
        !!glue("Min {pretty_label(input$xaxis_choose)}")  := min(.data[[input$xaxis_choose]]),
        !!glue("Mean {pretty_label(input$xaxis_choose)}") := mean(.data[[input$xaxis_choose]]),
        !!glue("Max {pretty_label(input$xaxis_choose)}")  := max(.data[[input$xaxis_choose]]))
  })
}

shinyApp(ui, server)

```

